---
title: "TR Location Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Setup of packages and functions
library(tidyverse)

source("src/source_all.R")
source_all("matteo_functions/")
source_all("src/")
```

# Introduction
This document describes an analysis of the location of various TR types in proteins associated with signalling pathways commonly implicated in CRC. The five main pathways analysed were: WNT, P53, PI3K, TGF-beta and RAS-MAPK. 
Proteins belonging to each pathway were obtained by querying UniProt for reviewed human proteins associated with a certain pathway. The set of proteins for each pathway was supplemented using lists of curated genes that are often affected in cancer. These lists were taken from Sanchez-Vega (2018) https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6070353/. For each gene, the encoded proteins were queried from UniProt and merged with the previously obtained set of proteins for each pathway. This resulted in the following number of proteins per pathway: WNT (698), P53 (701), PI3K (311), TGF-Beta (264), RAS-MAPK (126). In addition, proteins belonging to the DNA mismatch-repair (MMR) system were analysed (26 proteins, TODO), as well as proteins associated with favourable (404) or unfavourable (286) CRC prognosis in the Human Protein Atlas https://www.proteinatlas.org/humanproteome/pathology/colorectal+cancer.

# Results
```{r, include=FALSE}
# Important paths
base_path = "/Users/maxverbiest/PhD/projects/SP_CRC_Pathway_TRs/"
dest_file_sp <- "../data/swissrepeats/swissprot_human.tsv"
```

Below are two plots for each analysed pathway. One of the plots shows the distribution of all types of TRs (homo, micro, small, domain) in the pathway, the other shows the distribution of only the homo and small TRs. At the end of the document is an overview table showing some basic numbers for each pathway (how many proteins, how many TRs etc.).
### WNT pathway
```{r, include=FALSE}
# Load data: Protein source, TRs
wnt_data <- load_pathway_data(base_path, "wnt", meta = TRUE, idr = TRUE)
wnt_trs <- wnt_data["trs"][[1]]
wnt_meta <- wnt_data["meta"][[1]]
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
wnt_sp <- load_swissprot(dest_file_sp, wnt_trs)
wnt_trs_sp <- merge(x = wnt_trs, y = wnt_sp, by = "ID", all.x = TRUE)
wnt_trs_sp_meta <- merge(x = wnt_trs_sp, y = wnt_meta, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
summary_table_df <- data.frame(matrix(ncol=6,nrow=8, dimnames=list(c("wnt", "p53", "pi3k", "ras-mapk", "tgf-beta", "mmr", "pa_fav", "pa_unfav"), c("total_prots", "prots_with_TR", "total_TRs", "prots_with_TR_filt", "total_TRs_filt", "TR_prots_from_SV"))))

summary_table_df["total_prots"] <- c(698, 701, 311, 264, 126, 26, 404, 286)

# Total number of proteins containing TRs (before filter)
summary_table_df["wnt", "prots_with_TR"] <- length(unique(wnt_trs_sp_meta$ID))
# Total number of TRs (before filter)
summary_table_df["wnt", "total_TRs"] <- length(wnt_trs_sp_meta$ID)

# Filter for divergence and p-value of TRs
wnt_trs_sp_meta_filt = wnt_trs_sp_meta %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["wnt", "prots_with_TR_filt"] <- length(unique(wnt_trs_sp_meta_filt$ID))
# Total number of TRs after filter
summary_table_df["wnt", "total_TRs_filt"] <- length(wnt_trs_sp_meta_filt$ID)

# Count how many of the proteins containing TRs originate from the Sanchez-Vega list of curated genes
sv_count = 0
for (i in unique(wnt_trs_sp_meta_filt$ID)) {
  if (dim(wnt_trs_sp_meta_filt %>% filter(ID == i & SV == "yes"))[1] > 0) {sv_count = sv_count + 1}
}
summary_table_df["wnt", "TR_prots_from_SV"] <- sv_count
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  wnt_trs_sp_meta_filt,
  plot_title = "CRC WNT Protein TRs",
  byTRtype = TRUE)

wnt_trs_sp_meta_filt %>% 
  filter(l_type == "homo" | l_type == "micro") %>% 
  TR_location(plot_title = "CRC WNT homo and micro TRs", byTRtype = TRUE)
```

### P53 pathway
```{r, include=FALSE}
# Load data: Protein source, TRs
p53_data <- load_pathway_data(base_path, "p53", meta = TRUE, idr = TRUE)
p53_trs <- p53_data["trs"][[1]]
p53_meta <- p53_data["meta"][[1]]
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
p53_sp <- load_swissprot(dest_file_sp, p53_trs)
p53_trs_sp <- merge(x = p53_trs, y = p53_sp, by = "ID", all.x = TRUE)
p53_trs_sp_meta <- merge(x = p53_trs_sp, y = p53_meta, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
# Total number of proteins containing TRs (before filter)
summary_table_df["p53", "prots_with_TR"] <- length(unique(p53_trs_sp_meta$ID))
# Total number of TRs (before filter)
summary_table_df["p53", "total_TRs"] <- length(p53_trs_sp_meta$ID)

# Filter for divergence and p-value of TRs
p53_trs_sp_meta_filt = p53_trs_sp_meta %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["p53", "prots_with_TR_filt"] <- length(unique(p53_trs_sp_meta_filt$ID))
# Total number of TRs after filter
summary_table_df["p53", "total_TRs_filt"] <- length(p53_trs_sp_meta_filt$ID)

# Count how many of the proteins containing TRs originate from the Sanchez-Vega list of curated genes
sv_count = 0
for (i in unique(p53_trs_sp_meta_filt$ID)) {
  if (dim(p53_trs_sp_meta_filt %>% filter(ID == i & SV == "yes"))[1] > 0) {sv_count = sv_count + 1}
}
summary_table_df["p53", "TR_prots_from_SV"] <- sv_count
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  p53_trs_sp_meta_filt,
  plot_title = "CRC p53 Protein TRs",
  byTRtype = TRUE)

p53_trs_sp_meta_filt %>% 
  filter(l_type == "homo" | l_type == "micro") %>% 
  TR_location(plot_title = "CRC p53 homo and micro TRs", byTRtype = TRUE)
```

### PI3K pathway
```{r, include=FALSE}
# Load data: Protein source, TRs
pi3k_data <- load_pathway_data(base_path, "pi3k", meta = TRUE, idr = TRUE)
pi3k_trs <- pi3k_data["trs"][[1]]
pi3k_meta <- pi3k_data["meta"][[1]]
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
pi3k_sp <- load_swissprot(dest_file_sp, pi3k_trs)
pi3k_trs_sp <- merge(x = pi3k_trs, y = pi3k_sp, by = "ID", all.x = TRUE)
pi3k_trs_sp_meta <- merge(x = pi3k_trs_sp, y = pi3k_meta, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
# Total number of proteins containing TRs (before filter)
summary_table_df["pi3k", "prots_with_TR"] <- length(unique(pi3k_trs_sp_meta$ID))
# Total number of TRs (before filter)
summary_table_df["pi3k", "total_TRs"] <- length(pi3k_trs_sp_meta$ID)

# Filter for divergence and p-value of TRs
pi3k_trs_sp_meta_filt = pi3k_trs_sp_meta %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["pi3k", "prots_with_TR_filt"] <- length(unique(pi3k_trs_sp_meta_filt$ID))
# Total number of TRs after filter
summary_table_df["pi3k", "total_TRs_filt"] <- length(pi3k_trs_sp_meta_filt$ID)

# Count how many of the proteins containing TRs originate from the Sanchez-Vega list of curated genes
sv_count = 0
for (i in unique(pi3k_trs_sp_meta_filt$ID)) {
  if (dim(pi3k_trs_sp_meta_filt %>% filter(ID == i & SV == "yes"))[1] > 0) {sv_count = sv_count + 1}
}
summary_table_df["pi3k", "TR_prots_from_SV"] <- sv_count
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  pi3k_trs_sp_meta_filt,
  plot_title = "CRC pi3k Protein TRs",
  byTRtype = TRUE)

pi3k_trs_sp_meta_filt %>% 
  filter(l_type == "homo" | l_type == "micro") %>% 
  TR_location(plot_title = "CRC pi3k homo and micro TRs", byTRtype = TRUE)
```


### RAS-MAPK pathway
```{r, include=FALSE}
# Load data: Protein source, TRs
rasmapk_data <- load_pathway_data(base_path, "ras-mapk", meta = TRUE, idr = TRUE)
rasmapk_trs <- rasmapk_data["trs"][[1]]
rasmapk_meta <- rasmapk_data["meta"][[1]]
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
rasmapk_sp <- load_swissprot(dest_file_sp, rasmapk_trs)
rasmapk_trs_sp <- merge(x = rasmapk_trs, y = rasmapk_sp, by = "ID", all.x = TRUE)
rasmapk_trs_sp_meta <- merge(x = rasmapk_trs_sp, y = rasmapk_meta, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
# Total number of proteins containing TRs (before filter)
summary_table_df["ras-mapk", "prots_with_TR"] <- length(unique(rasmapk_trs_sp_meta$ID))
# Total number of TRs (before filter)
summary_table_df["ras-mapk", "total_TRs"] <- length(rasmapk_trs_sp_meta$ID)

# Filter for divergence and p-value of TRs
rasmapk_trs_sp_meta_filt = rasmapk_trs_sp_meta %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["ras-mapk", "prots_with_TR_filt"] <- length(unique(rasmapk_trs_sp_meta_filt$ID))
# Total number of TRs after filter
summary_table_df["ras-mapk", "total_TRs_filt"] <- length(rasmapk_trs_sp_meta_filt$ID)

# Count how many of the proteins containing TRs originate from the Sanchez-Vega list of curated genes
sv_count = 0
for (i in unique(rasmapk_trs_sp_meta_filt$ID)) {
  if (dim(rasmapk_trs_sp_meta_filt %>% filter(ID == i & SV == "yes"))[1] > 0) {sv_count = sv_count + 1}
}
summary_table_df["ras-mapk", "TR_prots_from_SV"] <- sv_count
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  rasmapk_trs_sp_meta_filt,
  plot_title = "CRC rasmapk Protein TRs",
  byTRtype = TRUE)

rasmapk_trs_sp_meta_filt %>% 
  filter(l_type == "homo" | l_type == "micro") %>% 
  TR_location(plot_title = "CRC rasmapk homo and micro TRs", byTRtype = TRUE)
```


### TGF-beta pathway
```{r, include=FALSE}
# Load data: Protein source, TRs
tgfb_data <- load_pathway_data(base_path, "tgf-beta", meta = TRUE, idr = TRUE)
tgfb_trs <- tgfb_data["trs"][[1]]
tgfb_meta <- tgfb_data["meta"][[1]]
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
tgfb_sp <- load_swissprot(dest_file_sp, tgfb_trs)
tgfb_trs_sp <- merge(x = tgfb_trs, y = tgfb_sp, by = "ID", all.x = TRUE)
tgfb_trs_sp_meta <- merge(x = tgfb_trs_sp, y = tgfb_meta, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
# Total number of proteins containing TRs (before filter)
summary_table_df["tgf-beta", "prots_with_TR"] <- length(unique(tgfb_trs_sp_meta$ID))
# Total number of TRs (before filter)
summary_table_df["tgf-beta", "total_TRs"] <- length(tgfb_trs_sp_meta$ID)

# Filter for divergence and p-value of TRs
tgfb_trs_sp_meta_filt = tgfb_trs_sp_meta %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["tgf-beta", "prots_with_TR_filt"] <- length(unique(tgfb_trs_sp_meta_filt$ID))
# Total number of TRs after filter
summary_table_df["tgf-beta", "total_TRs_filt"] <- length(tgfb_trs_sp_meta_filt$ID)

# Count how many of the proteins containing TRs originate from the Sanchez-Vega list of curated genes
sv_count = 0
for (i in unique(tgfb_trs_sp_meta_filt$ID)) {
  if (dim(tgfb_trs_sp_meta_filt %>% filter(ID == i & SV == "yes"))[1] > 0) {sv_count = sv_count + 1}
}
summary_table_df["tgf-beta", "TR_prots_from_SV"] <- sv_count
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  tgfb_trs_sp_meta_filt,
  plot_title = "CRC tgfb Protein TRs",
  byTRtype = TRUE)

tgfb_trs_sp_meta_filt %>% 
  filter(l_type == "homo" | l_type == "micro") %>% 
  TR_location(plot_title = "CRC tgfb homo and micro TRs", byTRtype = TRUE)
```


### DNA MMR
```{r, include=FALSE}
# Load data: Protein source, TRs
mmr_trs <- load_tr_annotations(paste0(base_path, "results/TRs/mmr_TRs.tsv"))
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
mmr_sp <- load_swissprot(dest_file_sp, mmr_trs)
mmr_trs_sp <- merge(x = mmr_trs, y = mmr_sp, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
# Total number of proteins containing TRs (before filter)
summary_table_df["mmr", "prots_with_TR"] <- length(unique(mmr_trs_sp$ID))
# Total number of TRs (before filter)
summary_table_df["mmr", "total_TRs"] <- length(mmr_trs_sp$ID)

# Filter for divergence and p-value of TRs
mmr_trs_sp_filt = mmr_trs_sp %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["mmr", "prots_with_TR_filt"] <- length(unique(mmr_trs_sp_filt$ID))
# Total number of TRs after filter
summary_table_df["mmr", "total_TRs_filt"] <- length(mmr_trs_sp_filt$ID)
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  mmr_trs_sp_filt,
  plot_title = "DNA MMR Protein TRs",
  byTRtype = TRUE)

mmr_trs_sp_filt %>% 
  filter(l_type == "homo" | l_type == "micro") %>% 
  TR_location(plot_title = "DNA MMR homo and micro TRs", byTRtype = TRUE)
```


### CRC favourable proteins
```{r, include=FALSE}
# Load data: Protein source, TRs
pa_fav_trs <- load_tr_annotations(paste0(base_path, "results/TRs/pa_fav_TRs.tsv"))
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
pa_fav_sp <- load_swissprot(dest_file_sp, pa_fav_trs)
pa_fav_trs_sp <- merge(x = pa_fav_trs, y = pa_fav_sp, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
# Total number of proteins containing TRs (before filter)
summary_table_df["pa_fav", "prots_with_TR"] <- length(unique(pa_fav_trs_sp$ID))
# Total number of TRs (before filter)
summary_table_df["pa_fav", "total_TRs"] <- length(pa_fav_trs_sp$ID)

# Filter for divergence and p-value of TRs
pa_fav_trs_sp_filt = pa_fav_trs_sp %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["pa_fav", "prots_with_TR_filt"] <- length(unique(pa_fav_trs_sp_filt$ID))
# Total number of TRs after filter
summary_table_df["pa_fav", "total_TRs_filt"] <- length(pa_fav_trs_sp_filt$ID)
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  pa_fav_trs_sp_filt,
  plot_title = "CRC pa_fav Protein TRs",
  byTRtype = TRUE)

pa_fav_trs_sp_filt %>% 
  filter(tr_type == "homo" | tr_type == "micro") %>% 
  TR_location(plot_title = "CRC pa_fav homo and micro TRs", byTRtype = TRUE)
```

### CRC unfavourable proteins
```{r, include=FALSE}
# Load data: Protein source, TRs
pa_unfav_trs <- load_tr_annotations(paste0(base_path, "results/TRs/pa_unfav_TRs.tsv"))
```


```{r, include=FALSE}
# Join with SwissProt information on proteins
pa_unfav_sp <- load_swissprot(dest_file_sp, pa_unfav_trs)
pa_unfav_trs_sp <- merge(x = pa_unfav_trs, y = pa_unfav_sp, by = "ID", all.x = TRUE)
```


```{r, include=FALSE}
# Total number of proteins containing TRs (before filter)
summary_table_df["pa_unfav", "prots_with_TR"] <- length(unique(pa_unfav_trs_sp$ID))
# Total number of TRs (before filter)
summary_table_df["pa_unfav", "total_TRs"] <- length(pa_unfav_trs_sp$ID)

# Filter for divergence and p-value of TRs
pa_unfav_trs_sp_filt = pa_unfav_trs_sp %>% 
  filter(divergence <= 0.1 & pvalue <= 0.05)
# Number of proteins containing TRs after filter
summary_table_df["pa_unfav", "prots_with_TR_filt"] <- length(unique(pa_unfav_trs_sp_filt$ID))
# Total number of TRs after filter
summary_table_df["pa_unfav", "total_TRs_filt"] <- length(pa_unfav_trs_sp_filt$ID)
```


```{r, warning=FALSE, echo=FALSE}
TR_location(
  pa_unfav_trs_sp_filt,
  plot_title = "CRC pa_unfav Protein TRs",
  byTRtype = TRUE)

pa_unfav_trs_sp_filt %>% 
  filter(tr_type == "homo" | tr_type == "micro") %>% 
  TR_location(plot_title = "CRC pa_unfav homo and micro TRs", byTRtype = TRUE)
```


```{r, echo=FALSE}
knitr::kable(summary_table_df)
```